---
title: "USFirearm"
author: "Syed Baryalay"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, fig.width=6, fig.height=3}
knitr::opts_chunk$set(echo = TRUE)
```


###

Generate some insightful visualizations to display this data. 

Does the rate of change in total firearms background checks over time vary across states?



```{r}
suppressMessages(library(tidyverse))
suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(fpp3))
suppressMessages(library(readr))
suppressMessages(library(forecast))
suppressMessages(library(ggplot2))

url <- "https://raw.githubusercontent.com/BuzzFeedNews/nics-firearm-background-checks/master/data/nics-firearm-background-checks.csv"

firearm <- read.csv(url, header = TRUE)
glimpse(firearm)
```



**Cleaning and filtering original data for analysis**


```{r}
library(dplyr)
library(lubridate)

dat <- firearm %>% 
    mutate(date = ymd(paste(month, "01"))) %>%
    select(date, state, totals) %>%
    filter(complete.cases(.)) %>%
    arrange(date) 

glimpse(dat)
```


**Plot of total background checks in each state as well as overall mean.**

```{r fig.width=6, fig.height=6}
ggplot(dat) + 
    geom_line(mapping=aes(x=date, y=totals, color=state)) +
    geom_line(dat %>% group_by(date) %>% summarise(totals=mean(totals)), 
              mapping=aes(x=date, y=totals)) +
    theme(legend.position="bottom",
          legend.title = element_text(size=10), 
          legend.text=element_text(size=6))
```


**Plot of total background checks in each state as well as overall median on the _log_ scale**

```{r fig.width=6, fig.height=6}
ggplot(dat) + 
    geom_line(mapping=aes(x=date, y=log(totals), color=state)) +
    geom_line(dat %>% group_by(date) %>% summarise(totals=median(totals)), 
              mapping=aes(x=date, y=log(totals))) +
    theme(legend.position="bottom",
          legend.title = element_text( size=10), 
          legend.text=element_text(size=6))
```



**Time series decomposition of total background firearm checks**

_we are only interested in trend which is obtain after removing seasonal effect_

```{r}
dat <- dat |> 
    group_by(date) |> 
    summarise(totals = sum(totals))

glimpse(dat) # totals checks each month in the US

firearm_ts <- ts(dat$totals, 
             start = c(min(year(dat$date)), 11), 
             end = c(max(year(dat$date)), 3), 
             frequency = 12)

# Decomposing the model to obtain trend
fit_dcmp <- decompose(firearm_ts)

autoplot(fit_dcmp) + 
  xlab("Year") + 
  ylab("Total Background Checks") + 
  ggtitle("USFirearm Decomposition")

autoplot(fit_dcmp$trend) + 
    xlab("Year") + 
    ylab("Total Background Checks") + 
    ggtitle("Trend of firearm background checks across states")
```


The original data has strong seasonality (no cycles however).

*After decomposing the time series, we have obtain the trend across the states.*


Next I fit an ARIMA model that I can use to forecast ahead.

```{r}
# fitting an ARIMA models to our time series
fit_arima <- auto.arima(firearm_ts)
summary(fit_arima)

# forecasting for the next 24 months
forecast_arima <- forecast(fit_arima, h = 24)

autoplot(forecast_arima) + 
  xlab("Year") + 
  ylab("Total Background Checks") + 
  ggtitle("24 Months Firearm Background Checks Forecast (ARIMA) ")
```